---
title: "Stats 306 Final Project"
author: "Stats 306 Instructional Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(dplyr)


```

## Investigating the Internet Movie Database (IMDB)

The [Internet Movie Database (IMDb)]() contains information on millions of movies and television programs. They offer several [non-commercial use datasets](https://developer.imdb.com/non-commercial-datasets/) (documentation link). For this project we will analyze a **sample** of 100,000 titles from the IMDBb. 


## Part I: Preprocessing

* [Edit your `.gitignore` file](https://docs.github.com/en/get-started/getting-started-with-git/ignoring-files) to ignore all files with the `.rda` extension. (Add and commit)
* Create a new file in the `data/` directory called "Preprocessing.Rmd". The remaining instructions in this section are to be completed in that file.
* Write a function that will load a table from the IMDb files in the `data/` directory.
  * The function should take the file name (without the ".csv.gz" portion) as an argument
  * The function should load the appropriate `.csv.gz` file.
  * Make sure that all "\\N" values (which IMDB uses to indicate missing values) are turned into proper NA values in R
  * Write the resulting table to a file called "FILENAME.rda" (where FILENAME is the argument to the function) using the `save()` function.
* Run the function on all of the `*_sample.csv.gz` files to created processed `.rda` files. You can load these using the `load()` function.


```{r}

str(name_basics) # chr - primaryName, primaryProfession
str(title_basics) # chr - titleType, genres, runtimeMinutes
str(title_principals) # chr -  category, job, characters
str(title_ratings) # chr - 

```


## Part II: EDA of individual tables

* For each of the 4 tables, perform basic exploratory data analysis. Report the following information:
  * For each quantitative column, provide some summary statistics
  * For any character columns, decided if they are actually representing factors/categorical data with a moderate number of columns. If so report the distributions for these variables.
  * Provide a plot for each table. Across all of the plots, try to show off the most possible different ggplot features (`geoms_` functions, `stat_` functions, coordinate systems, facets, use of several variables, annotations)
* For the `titles_basics` table
  * use two different variables to group and explore how `runtimeMinutes` varies for these different groups. Produce appropriate summaries.
  * How many titles are known for name that is different than the original release name?
  * Graph the conditional distributions of release year based on the previous results. Comment on any trends you observe.
* For the ratings, use the `cut` function to break the data into three groups based on the average ratings. Are higher rated titles rated more often or less often than lower rated titles? 


* For the names table, 
  * Count the number of titles each person is known for and plot this distribution.
  * investigate the age of cast members
      * Group the data into living and deceased cast members. 
      * For deceased cast members, provide a graph that shows the distribution of ages.
      * Do the same for living cast members.
* Find all the actors with first names "Tom", "Thomas", "Thom" or "Tomas". How many are there?
* How many titles use alliteration (i.e., all words in the title start with the same letter)?


```{r}
##  quantitative columns summary statistics

calculate_summary = function(column) {
  summary_stats = summary(column, na.rm = TRUE)
  sd_value = sd(column, na.rm = TRUE)
  
  summary_stats = c(summary_stats, SD = sd_value)
  return(summary_stats)
}

name_basics_stats = list(birthYear = calculate_summary(name_basics$birthYear), deathYear = calculate_summary(name_basics$deathYear))

title_basics_stats = list(isAdult = calculate_summary(title_basics$isAdult), startYear = calculate_summary(title_basics$startYear), endYear = calculate_summary(title_basics$endYear))

title_principals_stats = list(ordering = calculate_summary(title_principals$ordering))

title_ratings_stats = list(numVotes = calculate_summary(title_ratings$numVotes), averageRating = calculate_summary(title_ratings$averageRating))

all_summary_stats = list(name_basics = name_basics_stats, title_basics = title_basics_stats, title_principals = title_principals_stats, title_ratings = title_ratings_stats)

print(all_summary_stats)

```

```{r}
## categorical distributions

report_categorical_distribution <- function(column, column_name) {

  factor_column <- factor(column)

    cat("Distribution for", column_name, ":\n")
    print(table(factor_column))
    cat("\n")

}

report_categorical_distribution(name_basics$primaryProfession, "primaryProfession")

report_categorical_distribution(title_basics$titleType, "titleType")
report_categorical_distribution(title_basics$genres, "genres")

report_categorical_distribution(title_principals$category, "category")
report_categorical_distribution(title_principals$job, "job")

```


```{r}
## plots

# name basics
ggplot(name_basics, aes(x = birthYear)) +
  geom_histogram(binwidth = 10, fill = "orange", alpha = 0.7) +
  geom_histogram(aes(x = deathYear), binwidth = 10, fill = "blue", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Distribution of Birth and Death Years", x = "Year", y = "Count") +
  scale_x_continuous(limits = c(1800, 2020)) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Legend")) +
  scale_fill_manual(values = c("orange", "blue"), labels = c("Birth Year", "Death Year"))


# title basics
title_counts = title_basics |>
  group_by(startYear, titleType) |>
  summarise(count = n()) |>
  ungroup()

ggplot(title_counts, aes(x = startYear, y = titleType, fill = count)) +
  geom_tile() + 
  scale_fill_gradient(low = "lightblue", high = "blue") + 
  theme_minimal() +
  labs(title = "Heatmap of Title Types over the Years", x = "Start Year", y = "Title Type",fill = "Count") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 


# title principals
ggplot(title_principals, aes(x = ordering, fill = category)) +
  geom_density(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Density Plot of Ordering by Category", x = "Ordering", y = "Density") +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "right")


# title ratings
ggplot(title_ratings, aes(x = numVotes, y = averageRating)) +
  geom_point(aes(color = averageRating), alpha = 0.6) +
  scale_color_gradient(low = "blue", high = "red") +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  theme_minimal() +
  labs(title = "Scatter Plot of Average Rating vs Num Votes in Title Ratings", x = "Number of Votes", y = "Average Rating") +
  annotate("text", x = max(title_ratings$numVotes), y = min(title_ratings$averageRating), label = "Trend Line", hjust = 1, vjust = -1)

```
```{r}
# for Title basics

title_basics$runtimeMinutes = as.numeric(as.character(title_basics$runtimeMinutes))

title_basics_separated = title_basics |>
  separate_rows(genres, sep = ",") |>
  mutate(decade = cut(startYear, breaks = seq(1890, 2020, by = 10), labels = paste(seq(1890, 2010, by = 10), "s", sep = ""), include.lowest = TRUE))

runtime_summary_by_genre_and_year = title_basics_separated |>
  group_by(genres, decade) |>
  summarise(
    titles_count = n(),
    mean_runtime = mean(runtimeMinutes, na.rm = TRUE),
    median_runtime = median(runtimeMinutes, na.rm = TRUE),
    min_runtime = min(runtimeMinutes, na.rm = TRUE),
    max_runtime = max(runtimeMinutes, na.rm = TRUE),
    sd_runtime = sd(runtimeMinutes, na.rm = TRUE)
  ) |>
  ungroup() |>
  arrange(genres, decade)

print(runtime_summary_by_genre_and_year)

ggplot(runtime_summary_by_genre_and_year, aes(x = decade, y = mean_runtime, group = genres, color = genres)) +
  geom_line() +
  geom_point() + 
  theme_minimal() +
  labs(title = "Mean Runtime Minutes by Genre and Decade", x = "Decade", y = "Mean Runtime Minutes", color = "Genre") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) 


#

different_titles_count = title_basics |>
  filter(primaryTitle != originalTitle) |>
  summarise(count = n())

print(different_titles_count)
# How many titles are known for name that is different than the original release name? 7244

titles_with_different_names = title_basics |>
  filter(primaryTitle != originalTitle)

ggplot(titles_with_different_names, aes(x = startYear)) +
  geom_histogram(binwidth = 1, fill = "cornflowerblue") +
  theme_minimal() +
  labs(title = "Distribution of Release Years for Titles with Different Primary and Original Names", x = "Release Year", y = "Frequency")

# The most noticeable trend is the increasing number of titles with different primary and original names over time,
# especially in the second half of the 20th century and into the 21st century. This could suggest that the practice of
# using different names for the same title has become more common in recent times.

```

```{r}
# title ratings 

title_ratings = title_ratings |> 
  mutate(rating_group = cut(averageRating, breaks = quantile(averageRating, probs = c(0, 1/3, 2/3, 1), na.rm = T), labels = c("Low", "Medium", "High"), include.lowest = T))

rating_summary = title_ratings |>
  group_by(rating_group) |>
  summarise(mean_votes = mean(numVotes, na.rm = TRUE), total_num_votes = sum(numVotes, na.rm = TRUE), count = n())

print(rating_summary)

# it seems that for the higher ratings there are more votes than the rest of the rating groups. 

```

```{r}
# name basis

name_basics = name_basics |> 
  mutate(number_of_titles = sapply(strsplit(knownForTitles, ","), length))

head(name_basics$number_of_titles)

ggplot(name_basics, aes(x = number_of_titles)) +
  geom_histogram(binwidth = 1, fill = "coral4", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Number of Titles Each Person is Known For", x = "Number of Known For Titles", y = "Count of Persons")


#
name_basics = name_basics |>
  mutate(birthYear = as.numeric(birthYear), deathYear = as.numeric(deathYear), Age = ifelse(!is.na(deathYear), deathYear - birthYear, as.numeric(format(Sys.Date(), "%Y")) - birthYear))

living_cast = filter(name_basics, is.na(deathYear))
deceased_cast = filter(name_basics, !is.na(deathYear))


ggplot(deceased_cast, aes(x = Age)) +
  geom_histogram(binwidth = 1, fill = "tomato", color = "black") +
  theme_minimal() +
  labs(title = "Age Distribution of Deceased Cast Members", x = "Age at Death", y = "Count")


ggplot(living_cast, aes(x = Age)) +
  geom_histogram(binwidth = 1, fill = "dodgerblue", color = "black") +
  xlim(0, 130) +
  theme_minimal() +
  labs(title = "Age Distribution of Living Cast Members", x = "Current Age", y = "Count")


#
actors_named_tom = name_basics |>
  filter(str_detect(primaryName, "^Tom|^Thomas|^Thom|^Tomas"), str_detect(primaryProfession, "actor")) |>
  distinct(primaryName)

num_actors_named_tom = nrow(actors_named_tom)

num_actors_named_tom

# we have that 2051 actors are names Tom. 

```

```{r}
# titles 
is_alliterative = function(title) {

  words = unlist(str_split(title, "\\s+"))

  significant_words = words[!tolower(words) %in% c("the", "a", "an", "and", "but", "or", "for", "nor", "on", "at", "to", "from", "by")]
  if (length(significant_words) <= 1) {
    return(FALSE)  
  }
 
  first_letters = tolower(substr(significant_words, 1, 1))

  all(first_letters == first_letters[1])
}


alliterative_titles_count = title_basics |>
  filter(sapply(primaryTitle, is_alliterative)) |>
  summarise(count = n())

alliterative_titles_count

# we have that 3636 titles use alliteration. 

```


## Part III: Pivoting

* Create a new version of the `titles_basics` table that has one row for each title-genre combination.
* Using that table, create a line plot of the count different genres over time (you may limit this to the most common genres if you wish).
* Pivot the table to be a wide table with one column for each genre. Use this table to find the most common combination of genres.

## Part IV: Joining Tables

* Join the table with one title-genre per row from the previous section with the ratings table.
  * What is the highest rated genre? What is the lowest rated genre?
  * Using stacked bar charts, investigate the proportions of different genres over time. Are any incresing or decreasing? Use factor functions to help make the plots easier to read.
* Join the `title_basics` with the ratings table. Have the number of ratings changed over time (based on release year)? Display graphically but also answer with numerical results.
* Join the names with the ratings and the principals table. 
  * Group by individual people, find the top ten people based on the median rating of the titles they appear in.
  * Find the proportions of genres for the the titles that include the top 10 rated principals.
  * Graph ratings against years. What trends do you see?
* Create a table with one row for each person in the `name_basics` table and title they are known for. Join this to the ratings table to get the ratings of the "known for" films. Find the person (or people) who have the highest median known for rating.
* 

## Part V: Profiling and Parallel Processing

* These are large data sets (and yet only a sample of the entire IMDb!), so it make sense spend some time improving our code.
* Pick one or more of the previous problems and profile the performance of that piece. Write up your findings. If you see any opportunities to improve performance, feel fee to implement than and share the results.
* Select a previous computation that could be improved using parallelization and implement a parallelization solution. Using `system.time` show that parallelization improves performance.
* One task we performed involved counting items in strings separated by commas. Propose two different functions that could perform this taks. Compare them using bench marking. Which version would you recommend?

## Part VI: Shiny Applications

### Application 1

Using results from the previous section, create a shiny application that allows users to interact with the with the IMDb data. The application should use both interactive graphs and at least 3 widgets.

### Application 2

In the principals table, there is a `category` column. Use this column as a primary filter to allow users to then select specific job categories. After select the specific job categories, display information from another table.
